---
title: "GWAS Catalog store curation"
subtitle: "Delineating full-summary-statistic analyses into candidate OpenGWASDB stores"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    df-print: paged
    embed-resources: true
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup}
library(data.table)
library(ggplot2)

derived_dir <- "../data/derived"
summary_path <- file.path(derived_dir, "store-candidates-summary.tsv")
analyses_path <- file.path(derived_dir, "store-candidates-analyses.tsv")

# The curation itself lives in ebi-studies.r and writes the two TSVs below.
# Parsing the free-text sample-size field over the whole catalogue takes ~1
# minute, so this report reads the pre-computed outputs rather than re-running
# the pipeline. Regenerate them with:  Rscript resources/scripts/ebi-studies.r
if (!file.exists(summary_path) || !file.exists(analyses_path)) {
  stop("Derived tables not found. Run `Rscript resources/scripts/ebi-studies.r` first.")
}

store_summary <- fread(summary_path, sep = "\t", na.strings = "")
analyses <- fread(analyses_path, sep = "\t", na.strings = "")

fmt_int <- function(x) formatC(x, format = "d", big.mark = ",")
```

## What this is

Every row of the GWAS Catalog studies file is one **Analysis** (one trait
tested against variants). This report groups the analyses that ship **full
summary statistics** into candidate stores of three kinds:

- **dense** — one publication, one ancestry, ≥100 mostly complex/clinical-trait
  analyses from a shared cohort;
- **ragged** — one publication, one ancestry, ≥100 mostly molecular-trait
  analyses (proteomics, metabolomics, …);
- **hybrid** — everything left over, pooled by major ancestry group across many
  publications.

```{r headline}
n_full <- nrow(analyses)
by_type <- store_summary[, .(
  stores = .N,
  analyses = sum(n_analyses)
), by = store_type][order(-analyses)]

knitr::kable(
  rbind(by_type, data.table(store_type = "TOTAL",
                            stores = sum(by_type$stores),
                            analyses = sum(by_type$analyses))),
  col.names = c("Store type", "Candidate stores", "Analyses"),
  format.args = list(big.mark = ","),
  caption = sprintf("%s analyses with full summary statistics across %s candidate stores.",
                    fmt_int(n_full), fmt_int(nrow(store_summary)))
)
```

## Store size distribution

```{r size-dist, fig.width=8, fig.height=4.5}
ggplot(store_summary, aes(x = n_analyses, fill = store_type)) +
  geom_histogram(bins = 40, colour = "white", linewidth = 0.2) +
  scale_x_log10(labels = scales::comma) +
  facet_wrap(~store_type, scales = "free_y", ncol = 1) +
  labs(x = "Analyses per store (log scale)", y = "Number of stores",
       fill = "Store type") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

## Ragged stores by molecular type

Each ragged store is a single-publication molecular-QTL dataset. `molecular_type`
is taken from the study title (a store is one omics platform), falling back to
the dominant per-analysis trait subtype for method-titled papers.

```{r ragged-type-bar, fig.width=8, fig.height=4}
rag <- store_summary[store_type == "ragged"]
rag[, molecular_type := fifelse(is.na(molecular_type), "unresolved", molecular_type)]

type_counts <- rag[, .(stores = .N, analyses = sum(n_analyses)), by = molecular_type][order(-analyses)]

ggplot(type_counts, aes(x = reorder(molecular_type, analyses), y = analyses)) +
  geom_col(fill = "#4c72b0") +
  geom_text(aes(label = sprintf("%s stores", stores)), hjust = -0.1, size = 3.4) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.18))) +
  coord_flip() +
  labs(x = NULL, y = "Analyses across ragged stores") +
  theme_minimal(base_size = 12)
```

```{r ragged-type-table}
DT::datatable(
  rag[order(molecular_type, -n_analyses),
      .(store_key, molecular_type, ancestry = ancestry_group, n = n_analyses,
        pubmed_id, median_variants, median_sample_size,
        review = fifelse(is.na(review_reason), "", review_reason))],
  rownames = FALSE, filter = "top",
  options = list(pageLength = 15, scrollX = TRUE),
  caption = "Ragged stores labelled by molecular type.")
```

## Single-publication stores (dense & ragged)

Representative variant counts (`median_variants`) and sample sizes come from the
`PLATFORM [SNPS PASSING QC]` and `INITIAL SAMPLE SIZE` fields. `n_case_control`
/ `n_quantitative` split the analyses by study design. Dense stores are general
complex/clinical-trait batches (`molecular_type` empty); a dense store that does
carry a molecular type is flagged for review as a possible mis-split.

```{r dense-ragged-table}
sp <- store_summary[store_type != "hybrid"][order(-n_analyses)]
DT::datatable(
  sp[, .(store_key, type = store_type,
         molecular_type = fifelse(is.na(molecular_type), "", molecular_type),
         n = n_analyses, cc = n_case_control, quant = n_quantitative,
         med_variants = median_variants, med_N = median_sample_size,
         review = fifelse(is.na(review_reason), "", review_reason))],
  rownames = FALSE, filter = "top",
  options = list(pageLength = 15, scrollX = TRUE),
  caption = sprintf("All %d single-publication stores.", nrow(sp)))
```

### Variant count vs sample size

```{r variant-vs-n, fig.width=8, fig.height=5}
sp_pts <- copy(sp)
sp_pts[, fill_type := fifelse(store_type == "dense", "dense",
                       fifelse(is.na(molecular_type), "ragged (other)", molecular_type))]
ggplot(sp_pts[!is.na(median_variants) & !is.na(median_sample_size)],
       aes(x = median_sample_size, y = median_variants, colour = fill_type,
           size = n_analyses)) +
  geom_point(alpha = 0.75) +
  scale_x_log10(labels = scales::comma) +
  scale_y_log10(labels = scales::comma) +
  labs(x = "Median sample size per analysis (log)",
       y = "Median variants passing QC (log)",
       colour = "Store content", size = "Analyses") +
  theme_minimal(base_size = 12)
```

## Hybrid stores by ancestry

Left-over analyses pooled across publications, one store per major ancestry
group. Variant and sample-size medians here span many publications, so they are
indicative only.

```{r hybrid-table}
hy <- store_summary[store_type == "hybrid"][order(-n_analyses)]
knitr::kable(
  hy[, .(ancestry_group, n_analyses, n_pubmed_ids,
         median_variants, median_sample_size)],
  col.names = c("Ancestry group", "Analyses", "Publications",
                "Median variants", "Median N"),
  format.args = list(big.mark = ","))
```

## Analyses flagged for review

```{r review}
rv <- store_summary[needs_review == TRUE,
                    .(store_key, store_type, n_analyses, review_reason)][order(store_type, -n_analyses)]
cat(sprintf("%d of %d candidate stores flagged for manual review.\n\n",
            nrow(rv), nrow(store_summary)))
DT::datatable(rv, rownames = FALSE, filter = "top",
  colnames = c("Store", "Type", "Analyses", "Review reason"),
  options = list(pageLength = 10, scrollX = TRUE))
```

## Per-analysis sample sizes

Each analysis carries its own design and counts: case/control counts for binary
traits, total sample size for quantitative traits.

```{r per-analysis}
knitr::kable(
  analyses[, .(design = fifelse(is.na(study_design), "unknown", study_design),
               analyses = .N,
               median_N = as.integer(median(sample_size, na.rm = TRUE))),
           by = study_design][, .(design, analyses, median_N)][order(-analyses)],
  col.names = c("Study design", "Analyses", "Median sample size"),
  format.args = list(big.mark = ","))
```
