---
title: "Imputation quality filters for high-density stores"
subtitle: "Choosing MAF / imputation-r² cutoffs before importing TOPMed-imputed summary statistics"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    df-print: paged
    embed-resources: true
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
  cache: true
knitr:
  opts_chunk:
    cache.lazy: false
---

```{r setup}
#| cache: false
library(data.table)
library(ggplot2)
library(scales)

## Parameters ----------------------------------------------------------
## Point `accession` / `url` at any GWAS-SSF harmonised analysis to re-run.
## The MVP proteome store (dense__pmid-39024449__European) is the motivating
## case: ~44.3M variants per analysis, TOPMed-imputed.
accession <- "GCST90477290"
url <- paste0("http://ftp.ebi.ac.uk/pub/databases/gwas/summary_statistics/",
              "GCST90477001-GCST90478000/", accession,
              "/harmonised/", accession, ".h.tsv.gz")
data_dir <- "../data/mvp-example"
local_gz <- file.path(data_dir, basename(url))

## Store-size translation (matches ebi-studies.r).
n_analyses_store <- 3360     # analyses in the MVP European store
bytes_per_cell <- 2.5
gb <- function(n_variants) n_analyses_store * n_variants * bytes_per_cell / 1e9

fmt_int <- function(x) formatC(x, format = "d", big.mark = ",")
```

## Why this matters

The MVP store is imputed to ~44.3M variants per analysis. At 2.5 bytes/cell that
is `r round(gb(44.3e6))` GB for the store — and much of that tail is rare,
low-confidence imputed variants. This document looks at **one analysis** to
decide MAF / imputation-r² cutoffs before import. Because size scales linearly
with the retained variant count, the filter directly sets the store's cost and
credibility.

> Scope: a single analysis. Imputation r² varies across analyses, so confirm the
> chosen thresholds on a few more before locking them in — re-run this doc with a
> different `accession`.

```{r download}
#| cache: false
dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
if (!file.exists(local_gz)) {
  options(timeout = 7200)
  download.file(url, local_gz, mode = "wb", quiet = TRUE)
}
sprintf("%s: %.1f GB on disk", basename(local_gz), file.size(local_gz) / 1e9)
```

```{r load}
## Read only the two columns we need (fast, low memory on a ~10GB file), derive
## MAF, and bin it. Cached: this is the slow step; everything below is instant.
hdr <- names(fread(local_gz, nrows = 0))
freq_col <- intersect(c("effect_allele_frequency", "EAF", "eaf"), hdr)[1]
info_col <- intersect(c("r2", "info", "imputation_quality", "INFO", "IMPINFO"), hdr)[1]
stopifnot(!is.na(freq_col), !is.na(info_col))

dat <- fread(local_gz, select = c(freq_col, info_col))
setnames(dat, c(freq_col, info_col), c("eaf", "r2"))
dat <- dat[!is.na(eaf) & !is.na(r2)]
dat[, maf := pmin(eaf, 1 - eaf)]
dat <- dat[maf > 0 & maf <= 0.5]

maf_breaks <- c(0, 1e-4, 1e-3, 5e-3, 1e-2, 5e-2, 0.1, 0.2, 0.5)
maf_labels <- c("<0.0001", "0.0001–0.001", "0.001–0.005", "0.005–0.01",
                "0.01–0.05", "0.05–0.1", "0.1–0.2", "0.2–0.5")
dat[, maf_bin := cut(maf, breaks = maf_breaks, labels = maf_labels,
                     include.lowest = TRUE)]
n_total <- nrow(dat)
sprintf("%s variants with MAF and r2 (info column: '%s').", fmt_int(n_total), info_col)
```

## Distributions

```{r overview}
knitr::kable(
  rbind(
    data.table(metric = "MAF", t(round(quantile(dat$maf, c(0,.25,.5,.75,1)), 4))),
    data.table(metric = "r2",  t(round(quantile(dat$r2,  c(0,.25,.5,.75,1)), 3)))
  ),
  col.names = c("", "min", "Q1", "median", "Q3", "max"),
  caption = "Distribution of minor allele frequency and imputation r².")
```

```{r maf-dist, fig.width=8, fig.height=4}
mafd <- dat[, .(all = .N, pass = sum(r2 >= 0.8)), by = maf_bin][order(maf_bin)]
mafd <- melt(mafd, id.vars = "maf_bin", variable.name = "set", value.name = "n")
mafd[, set := factor(set, c("all", "pass"), c("all variants", "r² ≥ 0.8"))]
ggplot(mafd, aes(maf_bin, n, fill = set)) +
  geom_col(position = "dodge", width = 0.7) +
  scale_fill_manual(values = c("all variants" = "#9ec5f4", "r² ≥ 0.8" = "#2a78d6")) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  labs(x = "MAF bin", y = "Variants", fill = NULL,
       title = "Most variants sit in the rare bins — and most fail r² ≥ 0.8 there") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Imputation r² by MAF bin

The core question: rare variants impute worse, so a single r² cutoff is either too
lax for common variants or too harsh for rare ones. This is the evidence for a
**MAF-dependent** threshold.

```{r r2-by-maf-table}
byb <- dat[, .(
  n = .N,
  median_r2 = round(median(r2), 3),
  q25 = round(quantile(r2, .25), 3),
  q75 = round(quantile(r2, .75), 3),
  pct_pass_0.3 = round(100 * mean(r2 >= 0.3), 1),
  pct_pass_0.8 = round(100 * mean(r2 >= 0.8), 1)
), by = maf_bin][order(maf_bin)]
knitr::kable(byb, format.args = list(big.mark = ","),
  col.names = c("MAF bin", "n", "median r²", "r² Q1", "r² Q3",
                "% pass 0.3", "% pass 0.8"),
  caption = "Imputation r² collapses in the rare bins.")
```

```{r r2-by-maf-plot, fig.width=8, fig.height=4.5}
byplot <- dat[, .(median_r2 = median(r2),
                  q25 = quantile(r2, .25), q75 = quantile(r2, .75)), by = maf_bin][order(maf_bin)]
ggplot(byplot, aes(maf_bin, median_r2, group = 1)) +
  geom_ribbon(aes(ymin = q25, ymax = q75), fill = "#9ec5f4", alpha = 0.5) +
  geom_line(colour = "#2a78d6", linewidth = 0.9) +
  geom_point(colour = "#2a78d6", size = 2.4) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "#e34948") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "MAF bin", y = "Imputation r² (median, IQR band)",
       title = "Median imputation r² by MAF (dashed line = 0.8)") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Retention grid: what survives each (MAF, r²) cutoff

For every combination of a MAF floor and an r² floor, how many variants remain —
and what would the store cost if imported at that density?

```{r retention}
maf_thr <- c(0, 1e-4, 1e-3, 5e-3, 1e-2, 5e-2)
r2_thr <- c(0.3, 0.5, 0.7, 0.8, 0.9)
retention <- rbindlist(lapply(maf_thr, function(m) {
  r2v <- dat[maf >= m, r2]
  data.table(maf_min = m, r2_min = r2_thr,
             n_retained = vapply(r2_thr, function(r) sum(r2v >= r), numeric(1)))
}))
retention[, pct := round(100 * n_retained / n_total, 1)]
retention[, est_store_gb := round(gb(n_retained))]
```

```{r retention-heatmap, fig.width=8, fig.height=4.5}
ggplot(retention, aes(factor(r2_min), factor(maf_min), fill = pct)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = paste0(pct, "%\n", est_store_gb, "GB")), size = 3) +
  scale_fill_gradient(low = "#eef4fd", high = "#2a78d6", name = "% kept") +
  labs(x = "imputation r² floor", y = "MAF floor",
       title = "Variants retained (% of analysis) and implied store size",
       subtitle = paste0("Unfiltered: ", fmt_int(n_total), " variants ≈ ",
                         round(gb(n_total)), " GB for the ", fmt_int(n_analyses_store),
                         "-analysis store")) +
  theme_minimal(base_size = 12)
```

```{r retention-table}
knitr::kable(
  dcast(retention, maf_min ~ r2_min, value.var = "pct"),
  caption = "Percent of variants retained by MAF floor (rows) × r² floor (cols).")
```

## A tiered proposal

A single cutoff wastes either credibility or common-variant coverage. One
defensible tiered rule — **stricter r² where imputation is least reliable**:

- MAF ≥ 0.01 → r² ≥ 0.3
- 0.001 ≤ MAF < 0.01 → r² ≥ 0.8
- MAF < 0.001 → drop (imputation not trustworthy at this frequency)

```{r tiered}
dat[, keep_tiered :=
      (maf >= 0.01 & r2 >= 0.3) |
      (maf >= 0.001 & maf < 0.01 & r2 >= 0.8)]
tiered_n <- sum(dat$keep_tiered)

compare <- data.table(
  policy = c("unfiltered",
             "flat: MAF≥0.01 & r²≥0.3",
             "flat: MAF≥0.01 & r²≥0.8",
             "tiered (above)"),
  n_variants = c(n_total,
                 dat[maf >= 0.01 & r2 >= 0.3, .N],
                 dat[maf >= 0.01 & r2 >= 0.8, .N],
                 tiered_n)
)
compare[, pct_of_unfiltered := round(100 * n_variants / n_total, 1)]
compare[, est_store_gb := round(gb(n_variants))]
knitr::kable(compare, format.args = list(big.mark = ","),
  col.names = c("Policy", "Variants / analysis", "% of unfiltered", "Est. store GB"),
  caption = "Effect of candidate import policies on variant count and store size.")
```

## Takeaways

```{r takeaway}
#| cache: false
common_lax <- dat[maf >= 0.01 & r2 >= 0.3, .N]
cat(sprintf(
"- Unfiltered: %s variants/analysis -> ~%d GB store.\n",
  fmt_int(n_total), round(gb(n_total))))
cat(sprintf(
"- Common, lightly filtered (MAF>=0.01, r2>=0.3): %s variants (%.0f%%) -> ~%d GB.\n",
  fmt_int(common_lax), 100*common_lax/n_total, round(gb(common_lax))))
cat(sprintf(
"- Tiered policy: %s variants (%.0f%%) -> ~%d GB.\n",
  fmt_int(tiered_n), 100*tiered_n/n_total, round(gb(tiered_n))))
```

The rare tail is both the bulk of the variant count and the least trustworthy;
excluding MAF < 0.001 (or applying a strict r² there) removes most of the storage
and credibility risk while keeping essentially all common-variant signal. Confirm
the r²-by-MAF profile holds on a couple more analyses before setting a store-wide
policy.
